Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-17 manuscript/manuscript.Rmd
Minor edits

We report all sample sizes, data exclusions, manipulations, and
measures in the studies. Screen captures of the experimental materials
are available in the Supplemental Information. The complete data, code
to reproduce all statistical analyses and figures in the manuscript, as
well as the preregistration reports are available via OSF. All our
studies were preregistered on AsPredicted.org.^[<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>]

@@

# David: No need to redefine a factor variable as a continuous number. Easier to just change the reference level.

@@

However, recall that men were more confident in their performance than were women.
Our theory thus predicts that showing expectations should lead to more flattering
advice (i.e., a recommendation to compete against the High Performer group) for
men than for women. We report a linear probability model with advice to
compete against the high group as the outcome measure, and the advisee's
gender, whether expectations were shown to the adviser, and the
interaction of the two in Column 2 of Table \@ref(tab:study1regs). As
predicted, we find a significant interaction effect: men are more likely
to be advised to compete against the High Performers when expectations
are shown than when they are hidden ($p < 0.001$).
We show this result graphically in Figure \@ref(fig:study1genderdiff)). <!-- Amanda: The gender discrepancy in advice is more significant when performance is less decisive. When performance is extremely low, there is no chance to win high performers and only a slim chance when competing with low performers. When performance is extremely high, it makes no sense to forgo the high chance to win more and choose a low bonus. However, in the middle range of performance, either option has pros and cons, and neither is absolutely better. In such a situation, advisers would take participants' expectations into account, regarding them as a critical factor. -->

@@

@@

We recruited 300 participants from Prolific and, after asking demographic questions, invited them to upload photos of
themselves (selfies) to be rated by other participants on
attractiveness. 
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women agreed to do so and uploaded pictures that adhered to our
instructions (e.g., did not include other people). We selected the
first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years).
Participants were informed that their selfies would be randomly grouped with
those of nine other participants of their gender and ranked in terms of attractiveness
by a group of new Prolific participants of the opposite gender.
Lastly, they guessed their rank (unincentivized). **David: or did we start with those?**
<!-- Amanda: We started with demographic questions -->

Next, we recruited a new, gender-balanced sample for the role
of advisers. `r scales::comma(dim(study2_time2)[1])`
participants from Prolific ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started by
providing demographic information, then were matched to a group of the opposite
gender. They then ranked participants from most to least attractive using a
drop-down menu next to each picture. <!-- Amanda: We excluded data from participants who assigned the same rank to multiple selfies.-->
<!-- David: wasn't there something with participants selecting the same rank more -->
<!-- than once, and we dropped people who did that? Or were we able to solve this -->
<!-- and it happened only during a pilot? -->
<!-- Amanda: We did not catch this error in pilot, so we decided to use this as a manipulation check -->
After submitting their ratings, they saw the photo of the participant they
had ranked as the 7th most attractive (i.e., the 4th least attractive). They
were reminded of the rank they had given to that perso and informed
that this participant would be invited back and could earn a \$1 bonus if they
guessed their rank correctly. The rank was determined by the aggregate ratings
of all participants who had ranked this group. Their task was to give advice
to this participant about the rank they should bet on based on having observed
all ten selfies and their own assessment. We randomly assigned participants to
one of two incentivization schemes. In the "Accuracy" treatment, they received
a bonus identical to the advisee: \$1 if they guessed their rank correctly.
In the "Likeability" treatment, we informed them that they would be rated by
the advisee on a 5-point Likert scale for how likeable they thought they were.
Each point on the scale would translate to a bonus of 20 cents. They then
selected a rank that they would recommend the advisee to bet on.

@@

<!-- Amanda: How about this -->
```{r study2Rating, fig.cap="The evaluation of advice givers"}
# Load the necessary packages
library(ggplot2)
library(gridExtra)
library(dplyr)

@@

We were not powered to do a comparison across the two 
experimental groups and did not preregister such a difference. Indeed, we observed that advisers in the Accurate condition received higher ratings in both likable and warm dimensions (`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`, and `r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`, respectively). While the difference is not statistically significant, the results suggest that there may be an optimal range for the level of inflation.
<!-- David: But we still need to report warmth & trust across treatments. -->

@@

# General Discussion
Advice has the potential to shape people's career and personal outcomes.
Honest feedback, however, may be painful to learn if it falls short of one's
expectations. As prior work notes, this may motivate people to avoid information 
and avoid seeking help [@Benabou2022; @Golman2017; @Jaroszewicz2022].
We present evidence from two experiments that advisers are also cognizant of this
cost. As a result, they present flattering advice that avoids disappointing the
recipient, and correctly anticipate that this boosts how advisees perceive them.

However, this flattering advice comes at a cost to advisees, who would do worse
if they followed it blindly, as we show in Study 1.

<!-- David: But not Study 2? I guess we're not sure yet? -->

<!-- Amanda: among 146 participants in time 3, only 25 guessed correctly. not enough variation-->
<!-- Nonetheless, participants received advice from accurate condition earned less (0.15 vs 0.2) -->

>>>>

We report all sample sizes, data exclusions, manipulations, and
measures in the studies. Screen captures of the experimental materials
are available in the Supplemental Information. The complete data, code
to reproduce all statistical analyses and figures in the manuscript, as
well as the preregistration reports will be available via OSF. All our
studies were preregistered on AsPredicted.org.

<!-- ^[<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>] -->
<!-- Commenting this out for IACM -->

@@

@@

However, recall that men were more confident in their performance than were women.
Our theory thus predicts that showing expectations should lead to more flattering
advice (i.e., a recommendation to compete against the High Performer group) for
men than for women. We report a linear probability model with advice to
compete against the high group as the outcome measure, and the advisee's
gender, whether expectations were shown to the adviser, and the
interaction of the two in Column 2 of Table \@ref(tab:study1regs). As
predicted, we find a significant interaction effect: men are more likely
to be advised to compete against the High Performers when expectations
are shown than when they are hidden ($p < 0.001$).
We show this result graphically in Figure \@ref(fig:study1genderdiff)).

<!-- Amanda: The gender discrepancy in advice is more significant when performance is less decisive. When performance is extremely low, there is no chance to win high performers and only a slim chance when competing with low performers. When performance is extremely high, it makes no sense to forgo the high chance to win more and choose a low bonus. However, in the middle range of performance, either option has pros and cons, and neither is absolutely better. In such a situation, advisers would take participants' expectations into account, regarding them as a critical factor. -->

<!-- David: If we plotted on the x-axis performance and on the y-axis the difference -->
<!-- between recommendations for men & women, this would look inverse-U shaped, right? -->
<!-- Might be a nice way to illustrate what you describe. -->

@@

<!-- Yes, new table is fine. -->

@@

We recruited 300 participants from Prolific and, after asking demographic questions, invited them to upload photos of
themselves (selfies) to be rated by other participants on
attractiveness. 
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women agreed to do so and uploaded pictures that adhered to our
instructions (e.g., did not include other people). We selected the
first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years).
Participants were informed that their selfies would be randomly grouped with
those of nine other participants of their gender and ranked in terms of attractiveness
by a group of new Prolific participants of the opposite gender.
Lastly, they guessed their rank (unincentivized).

Next, we recruited a new, gender-balanced sample for the role
of advisers. `r scales::comma(dim(study2_time2)[1])`
participants from Prolific ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started by
providing demographic information, then were matched to a group of the opposite
gender. They then ranked participants from most to least attractive using a
drop-down menu next to each picture. Because of a limitation with the survey software,
participants were able to select the same rank for multiple participants, rather
than rank them uniquely. We remove XX participants who did not follow instructions
and provide a unique ranking.

<!-- David: Note the number above to fill in. -->

After submitting their ratings, they saw the photo of the participant they
had ranked as the 7th most attractive (i.e., the 4th least attractive). They
were reminded of the rank they had given to that perso and informed
that this participant would be invited back and could earn a \$1 bonus if they
guessed their rank correctly. The rank was determined by the aggregate ratings
of all participants who had ranked this group. Their task was to give advice
to this participant about the rank they should bet on based on having observed
all ten selfies and their own assessment. We randomly assigned participants to
one of two incentivization schemes. In the "Accuracy" treatment, they received
a bonus identical to the advisee: \$1 if they guessed their rank correctly.
In the "Likeability" treatment, we informed them that they would be rated by
the advisee on a 5-point Likert scale for how likeable they thought they were.
Each point on the scale would translate to a bonus of 20 cents. They then
selected a rank that they would recommend the advisee to bet on.

@@

<!-- Amanda: How about this -->
<!-- For warmth it looks interesting, but I think the plot is hard -->
<!-- to read. Let's hold off on this for now. -->

```{r study2Rating, fig.cap="The evaluation of advice givers", eval = F}
# Load the necessary packages
library(ggplot2)
library(gridExtra)
library(dplyr)

@@

We were not powered to do a comparison across the two 
experimental groups and did not preregister such a difference. Indeed, we find
no difference in likeability and warmth across the two treatments 
(`r papaja::apa_print(t.test(likable ~ Condition, data = study2_time3, var.equal = T))$statistic`, 
and `r papaja::apa_print(t.test(warmAvg ~ Condition, data = study2_time3, var.equal = T))$statistic`, 
respectively).

@@

# General Discussion
Advice has the potential to shape people's career and personal outcomes.
Honest feedback, however, may be painful to learn if it falls short of one's
expectations. As prior work notes, this may motivate people to avoid information 
and avoid seeking help [@Benabou2022; @Golman2017; @Jaroszewicz2022].
We present evidence from two experiments that advisers are also cognizant of this
cost. As a result, they present flattering advice that avoids disappointing the
recipient, and correctly anticipate that this boosts how advisees perceive them.
However, this flattering advice comes at a cost to advisees, who would do worse
if they followed it blindly, as we show in Study 1.
